---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RtoSQLServer

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R-CMD-check](https://github.com/DataScienceScotland/RtoSQLServer/workflows/R-CMD-check/badge.svg)](https://github.com/DataScienceScotland/RtoSQLServer/actions)
<!-- badges: end -->

R package used to import R dataframes into a MS SQL Server database, optionally with [System Versioning](https://docs.microsoft.com/en-us/sql/relational-databases/tables/creating-a-system-versioned-temporal-table?view=sql-server-ver15) enabled. 

## Installation

R package can be installed directly from Github or locally from zip.  
To install directly from GitHub:

```r
remotes::install_github(
  "DataScienceScotland/rtosqlserver",
  upgrade = "never"
)
```

If the above does not work, install by downloading:

1. Download the [zip of the repository](https://github.com/DataScienceScotland/sgplot/archive/refs/heads/main.zip) from GitHub.
2. Save the downloaded zip to a specific directory (e.g. C:/temp).
3. Install with this command specifying the path to the downloaded zip:
```
remotes::install_local("C:/temp/RtoSQLServer-main.zip", upgrade="never")
```

## Functionality

As well as loading R dataframes into SQL Server databases, functions are also currently available to: 

- Select all rows of specific columns of a database table.
- Drop a table from the database.
- Run any other input sql in the database and return dataframe if a select statement.  

It is recommend to ensure using the latest versions of [Rcpp](https://cran.r-project.org/web/packages/Rcpp/index.html), [odbc](https://cran.r-project.org/web/packages/odbc/index.html) and [DBI](https://cran.r-project.org/web/packages/DBI/index.html). If using Windows in a secure environment install these from source or a Windows binary compiled at the version of R you are using.

## Loading method used
When loading an R dataframe into SQL Server using `write_dataframe_to_db`, following steps are followed:

1. The R dataframe is loaded into a staging table in the database in batches of n rows at a time.

2. a) If table of the specified name does NOT already exist in the database schema:  
      i) Create target table in the database.  
      ii) Insert all rows from staging table to target table.

3. b) If table of same name does already exist in the database schema:  

    If 'append_to_existing'=FALSE (this will result in an overwrite):  
      i) Drop the existing copy of the target table and create a new one from staging table definition.  
      ii) Insert all rows from staging table into target table.  

    If 'append_to_existing'=TRUE:  
      i) Check that staging table columns and existing target table columns are the same. If not, cancel loading and give a warning.  
      ii) If check passes, insert all rows from staging table into target table.  

4. Delete the staging table.

## Example Usage
Here is an example using the main functions:
```{r eval=FALSE}
# Make a test dataframe with n rows
test_n_rows <- 1234567
test_df <- data.frame(a=rep("a", test_n_rows), b=rep("b", test_n_rows))

# Set database connection details for use in functions:
server <- "server\\instance"
database <- "my_database_name"
schema <- "my_schema_name"

# Write the test dataframe to a SQL Server table in 100K batches (by default system versioning is FALSE)
write_dataframe_to_db(server=server, 
                      database=database, 
                      schema=schema, 
                      table_name="test_r_tbl", 
                      dataframe=test_df, 
                      append_to_existing = FALSE,
                      batch_size=1e5, 
                      versioned_table=FALSE)

# Read the SQL Server table into an R dataframe
read_df <- read_table_from_db(server=server, 
                              database=database, 
                              schema=schema, 
                              table_name="test_r_tbl")
                              
# Run other SQL for example select specific column
sql <- paste0("select a from ", schema, ".test_r_tbl")

read_selected_df <- execute_sql(server=server,
                                database=database,
                                sql=sql,
                                output=TRUE)
                              

# Drop the table from the database
drop_table_from_db(server=server, 
                   database=database, 
                   schema=schema, 
                   table_name="test_r_tbl")
```
