---
title: "Read database table with column or row filtering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Read database table with column or row filtering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
The `read_table_from_db` function includes optional arguments so you can read only specific columns into an R dataframe or so you can filter which rows are read.  

To select only specific columns, specify a vector of column_names for example `c("Sepal_Length", "Species")` for the `columns` argument of `read_table_from_db`.  

To select only specific rows, specify a filter string using R syntax (not SQL syntax) for the `filter_stmt` argument. For example `"(Species == 'setosa' | Species == 'virginica') & Sepal_Length > 5.0"`. The function will convert this to SQL.

If you are unsure of column names in an existing database table, use the `db_table_metadata` function to check.

Here's a full example of both column and row selection using the `columns` and `filter_stmt` arguments of `read_table_from_db` for a copy of the iris data loaded into the database:

```{r eval=FALSE}
library(RtoSQLServer)

db_test_iris <- read_table_from_db(
  server = server,
  database = database,
  schema = schema,
  table_name = "test_iris",
  columns = c("Sepal_Length", "Species"),
  filter_stmt = "(Species == 'setosa' | Species == 'virginica') & Sepal_Length > 5.0",
  include_pk = FALSE
)
```
For the `filter_stmt` argument, this should be passed as a string. If you find the syntax confusing when wrapping your filter within quotation marks, then instead try using `deparse1(substitute())`. For example, the above filter example could be passed as `deparse1(substitute((Species == 'setosa' | Species == 'virginica') & Sepal_Length > 5.0))`.  

If you are comfortable with SQL, you may prefer to use the `execute_sql` function where a SQL select statement can be input to return an R dataframe from a database.
